Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-04-05 doc/manuscript_hkk_drought_sensitivity.Rmd
more edits to paragraphs 1 & 2, starting on 3

<!--KAT suggestion for new intro paragraph (would replace next paragraph)-->
In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@intergovernmentalpanelonclimatechangeipcc_climate_2023]. 
Understanding these impacts is an important research priority, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019].
However, study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
Separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [@bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought characteristics can influence climatic drivers of tree growth and their timing, and affect drought sensitivity of tropical trees. 
Droughts, or hot/dry conditions anomalously departing from their long-term means during any given time of year, are notoriously hard to define and encompass a variety of types based on the meteorogical or climatic variables affected [@slette_how_2019]. 
<!--Here we define drought as hot/dry conditions anomalously departing from their long-term means during any given time of year. -->
<!-- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  -->
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
In tropical seasonal forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically in relation to the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. 
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
Tree species in seasonally dry forests have diverse evolutionary strategies <!--that span and interact across timescales --->to maximise growth and survival under periodic hot/dry conditions. 
These strategies can range along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions[@gonzález-m._diverging_2021]. 
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid water loss and have acquisitive strategies during leaf on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf turgor loss point (hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), large vessels (hydraulic efficiency) [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls; e.g. some species regulate negative TLPs, leaf deciduousness and root depth, based on environmental cues leading to intraspecific or interannual variation potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky? Hulshof?].

>>>>

In forests worldwide, intense drought is becoming more frequent, affecting forest demography, diversity, and  C cycling [@intergovernmentalpanelonclimatechangeipcc_climate_2023]. 
Understanding these impacts is an important research priority, particularly for tropical forests, whose critical C sink is threatened by drought [@brando_droughts_2019].
However, study of tropical tree woody growth responses to drought is inhibited by the fact that few tropical species form rings and long-term annual records are rare [@zuidema_tropical_2022]. 
For forests worldwide, separate studies have shown that drought characteristics, species traits, and microenvironment all affect tree growth responses [e.g., @bennett_larger_2015;@mcgregor_tree_2021], but we know little about their combined effects or potential interactions, which could fundamentally alter our understanding of forest responses to drought. 
Specifically, if drought tolerance is influenced independently by drought characteristics, species traits, and microenvironment, tree drought responses may be conserved, and therefore linearly predictable; i.e., a species with low sensitivity to past droughts would have low sensitivity in any microhabitat and through any drought. 
In contrast, if these factors interact, the species and individuals that exhibit resistance to one drought may prove much more vulnerable in a different drought. 
The resultant amplification of the complexity and diversity of drought responses may act as a mechanism for promoting species diversity and stabilizing ecosystem function as drought regimes intensify [@dahinden_future_2017;@luo_tropical_2022;@naumann_global_2018].

@@

<!--PARAGRAPH ON DROUGHT CHARACTERISTICS-->
Drought, or hot/dry conditions anomalously departing from climatic means during any given time of year, is notoriously hard to define and encompasses a variety of types based on the meteorological variables affected and the severity and duration of the climatic anomalies [@slette_how_2019]. 
While forest drought responses are often considered to follow consistent patterns across droughts [e.g.,@bennett_larger_2015; @refs ], droughts with different characteristics may in fact have very different effects on tree growth [@refs]. 
<!--Here we define drought as hot/dry conditions anomalously departing from their long-term means during any given time of year. -->
<!-- Droughts are characterised by anomalously hot and/or dry conditions, but these can occur at any time during the seasonal phenology of tree growth.  -->
<!-- Tropical tree growth shows predictable seasonal patterns [@garcía-cervigón_climate_2020;@giraldo_tree_2023]; drought timing can interact with this growth phenology especially.  -->
In tropical seasonal forests with dry seasons unfavorable to growth, tree drought responses may vary dramatically in relation to the timing of drought relative to seasonal cycles of climate, photosynthesis, and woody growth [@garcía-cervigón_climate_2020]. 
<!--(Here we distinguish dry season from drought; dry season represents hot/dry conditions that are periodic, annual and predictable, droughts are unpredictable eith anomalous hot, dry conditions relative to seasonal expectations.) -->
Analyses to date have found that the woody growth of tropical trees <!--decreases with temperature[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021; @vlam_temperature_2014] and -->is particularly vulnerable to dry season precipitation[@zuidema_tropical_2022;@clark_annual_2010;@clark_annual_2021]; however, these analyses represent only a very limited set of tropical tree species and climates. 
It remains unknown how diverse assemblages of tree species in tropical dry forests respond to different types of droughts, including those occurring at different times relative to the regular seasonal cycle of wet and dry seasons. 
<!--
Seasonally dry tropical forest climates are often close to thermal and water stress tolerance thresholds for many species therein, making these forests vulnerable to drought [@allen_will_2017] and leading to stand-level patterns of drought-associated water stress [@bauman_tropical_2022; @bennett_sensitivity_2023; @aguirre-gutiérrez_drier_2019]. <!--the last phrase of this sentence and how it relates to Bennett et al. is unclear to me.
Trees in seasonally dry tropical forests exhibit a mix of strategies - along a spectrum of conservative to acquisitive water-use - that support survival and growth despite strong seasonal hot/dry stress.
These strategies might confer differential advantages under drought conditions. -->

<!--PARAGRAPH ON HOW SPECIES TRAITS (AND VARIATION) SHAPE DROUGHT RESPONSES-->
As in forests globally, tree species in seasonally dry forests have diverse evolutionary strategies <!--that span and interact across timescales --->to cope with drought, often leading to classification of species as "drought tolerant" or "drought sensitive" [@refs].
In tropical seasonal forests in particular, trees have evolved to maximise growth and survival under periodic hot/dry conditions, with strategies ranging along a spectrum from hydraulic safety to efficiency leading to tolerance or avoidance of dry conditions [@gonzález-m._diverging_2021]. 
Dry season deciduousness represents an extreme "avoidance" strategy along this continuum, where species lose leaves during the dry season to avoid water loss and have acquisitive strategies during leaf on periods [@desouza_drought_2020]. 
Other strategies include deep roots (hydraulic safety, to access deeper water in the dry season), more negative leaf turgor loss point (hydraulic safety, to maintain gas exchange and photosynthesis under hot and dry conditions), large vessels (hydraulic efficiency) [@gonzález-m._diverging_2021;@chitra-tarak_hydraulicallyvulnerable_2021]. 
Under drought conditions, some of these strategies can lend drought resistance while others can amplify drought sensitivity especially under severe droughts [@kunert_leaf_2021; @chitra-tarak_hydraulicallyvulnerable_2021; @gonzález-m._diverging_2021]. 
<!-- This spectrum of species strategies in seasonally dry forests for demographic stability under hot dry seasons might have distinct consequences for sensitivity, based on drought characteristics.  -->
<!--Some theoretical evidence that the relative advantage of evergreen and deciduous strategies might differ with changing precipitation regimes [@vico_how_2017a]--->
Moreover, besides being constant or adaptively conserved, these strategies can also have developmental or ecophysiological controls; e.g. some species regulate negative TLPs, leaf deciduousness and root depth, based on environmental cues leading to intraspecific or interannual variation potentially resulting in spatiotemporal variation in drought responses [@williams_deciduousness_2008; REFS Lasky? Hulshof?].
